Discussion

### 2020-08-03 code/figure4_function.Rmd

  gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
  arrange(as.numeric(order)) %>% .$gene_name ## getting the gene name order
  ## gene mutation data and tumor samples - Booleanize mutation count for each tumor sample
  message("working on mutation data")
  filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
    left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
    column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
    .[rownames(.) %in% gene_order, names(.) %in% class_spec] %>% ## filter by genes in top10 path of filtered class and tumor samples associated with it
    .[match(gene_order, rownames(.)),]
  message("Running pheatmap to get coordinates")
  pheatmap_res <- pheatmap(filtered_mut_data,
         show_rownames = T,
         show_colnames = F,
         treeheight_row = 0,
         treeheight_col = 0,
         cellheight = 5,
         cellwidth = 1,
         cluster_cols = T,
         cluster_rows = F, ## turn off row clustering
         annotation_legend = F,
         legend_breaks = c(0, 0.25, 0.75, 1),
         legend_labels = c("","Not Mut", "Mut", ""),
         color = c("grey", "firebrick3"),
         clustering_method = "ward.D2",
         fontsize = 10,
         legend = T)
  dev.off()
  pheatmap_order <- pheatmap_res$tree_col$order ## getting sample clustering order
  sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
  sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
  message("Preparing data for plot")
  gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
  gene_2_path_heatmap$membership <- as.character(gene_2_path_heatmap$membership) ## turn boolean values into categories
  ## set levels based on clustering algorithm above
  gene_2_path_heatmap <- left_join(gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
  gene_2_path_heatmap$gene_name <- factor(gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
  gene_2_path_heatmap$sample_id <- factor(gene_2_path_heatmap$sample_id, levels = c(sample_order))
  heatmap_ggplot <- ggplot(data = gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  xlab("Tumor Samples") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_blank(),
        strip.background = element_blank(),
        legend.position = "none") +
    ggtitle(paste0("Class ", filtered_class))
  ## left annotatoin data generation
  l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap

@@

  plotlist <- c(l_anno_ggplot, list(heatmap_ggplot)) ## make a list of plots for cowplot

  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 11, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 5))
  plot_legend <- plot_grid(plotlist = path_legend, ncol = 1, align = "v")

@@

```{r figure-4-plot, out.width= '90%', fig.cap="**Figure 4: Some genes with multiple pathway memberships drive clustering.** Each column is a tumor in class 6 or 7, each row represents a gene with mutation (red) or not mutated (grey). The genes are organized according to which of the top distinguishing pathways for this class that they effect. Some genes affect multiple pathways, as indicated by the bars at left. The highlighted genes at right are all proteasomal subunits."}
load("../data/class_spec_fig4.Rda")
class_spec_fig4
```

>>>>

  gene_order <- path_match_in_set %>% left_join(., most_genes, by = "path_combo") %>% ## left join the data together
  arrange(as.numeric(order)) %>% .$gene_name ## getting the gene name order
  ## gene mutation data and tumor samples - Booleanize mutation count for each tumor sample
  message("working on mutation data")
  filtered_mut_data <- gdc_mut_count_patient %>% rownames_to_column(var = "ensembl_gene_id") %>% ## ensembl id to col
    left_join(., gene_data[,c("ensembl_gene_id", "external_gene_name")], by = "ensembl_gene_id") %>% ## left_join data by ensembl_gene_id
    column_to_rownames(var = "external_gene_name") %>% dplyr::select(-ensembl_gene_id) %>% ## replace ensembl_id with gene_name
    .[rownames(.) %in% gene_order, names(.) %in% class_spec] %>% ## filter by genes in top10 path of filtered class and tumor samples associated with it
    .[match(gene_order, rownames(.)),]
  message("Running pheatmap to get coordinates")
  pheatmap_res <- pheatmap(filtered_mut_data,
         show_rownames = T,
         show_colnames = F,
         treeheight_row = 0,
         treeheight_col = 0,
         cellheight = 5,
         cellwidth = 1,
         cluster_cols = T,
         cluster_rows = F, ## turn off row clustering
         annotation_legend = F,
         legend_breaks = c(0, 0.25, 0.75, 1),
         legend_labels = c("","Not Mut", "Mut", ""),
         color = c("grey", "firebrick3"),
         clustering_method = "ward.D2",
         fontsize = 10,
         legend = T)
  dev.off()
  pheatmap_order <- pheatmap_res$tree_col$order ## getting sample clustering order
  sample_order <- names(filtered_mut_data) ## getting the sample_id original order 
  sample_order <- sample_order[pheatmap_order] ## reorder cancer by heatmap order
  message("Preparing data for plot")
  gene_2_path_heatmap <- filtered_mut_data %>% rownames_to_column(var = "gene_name") %>%  ## move rownames to column
  gather(., key = "sample_id", value = "membership", -gene_name) ## wide to long format
  gene_2_path_heatmap$membership <- as.character(gene_2_path_heatmap$membership) ## turn boolean values into categories
  ## set levels based on clustering algorithm above
  gene_2_path_heatmap <- left_join(gene_2_path_heatmap, path_match_in_set, by = "gene_name") ## add in path_combo for faceting 
  gene_2_path_heatmap$gene_name <- factor(gene_2_path_heatmap$gene_name, levels = c(rev(gene_order)))
  gene_2_path_heatmap$sample_id <- factor(gene_2_path_heatmap$sample_id, levels = c(sample_order))
  heatmap_ggplot <- ggplot(data = gene_2_path_heatmap, aes(y=gene_name, x=sample_id, fill = membership)) + 
  geom_tile(size = 0.6) +
  scale_fill_manual(values = c("grey", "firebrick3")) +
  xlab("Tumor Samples") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.ticks = element_blank(),
        strip.text.y = element_blank(),
        strip.background = element_blank(),
        plot.margin = unit(c(0,0,0,0), "cm"),
        legend.position = "none") +
    ggtitle(paste0("Class ", filtered_class))
  ## left annotatoin data generation
  l_anno <- path_match_in_set %>% .[match(gene_order, .$gene_name),] ## order by the gene order from the heatmap

@@

  proteasome_r_anno <- data.frame(gene = gene_order, 
                                psm_gene = 0)

  proteasome_r_anno[grep(pattern = "PSM|SEM|UBE2|UBB|UBC", x = proteasome_r_anno$gene), "psm_gene"] <- 1

  r_anno_color <- c("0" = "white", "1" = "black")
  proteasome_r_anno$gene <- factor(proteasome_r_anno$gene, levels = c(rev(gene_order)))
  r_anno <- ggplot(proteasome_r_anno) +
    geom_bar(mapping = aes(x = 1, y = gene, fill = as.character(psm_gene)), 
             stat = "identity", 
            width = 1) +
    scale_fill_manual(values = r_anno_color, breaks = "1", name = "", labels = "PSM Gene") +
    theme_void() 
  legend <- plot_grid(get_legend(r_anno), ncol = 1)
  r_anno <- r_anno + theme(legend.position = "none")

  plotlist <- c(l_anno_ggplot, list(heatmap_ggplot), list(r_anno)) ## make a list of plots for cowplot

  plot <- plot_grid(plotlist = plotlist, align = "h", ncol = 12, axis = "tb", rel_widths = c(rep(0.1, length(l_anno_ggplot)), 5, 0.1))
  plot_legend <- plot_grid(plotlist = path_legend, ncol = 1, align = "v")

@@

```{r figure-4-plot, out.width= '90%'}
load("../data/class_spec_fig4.Rda")
class_spec_fig4
```

<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Figure 3: Mutations in proteasomes found in Class 2 affect multiple enriched pathways.** a) Each column is a tumor in cluster 2, each row represents a gene with mutation (red) or not mutated (grey). The genes are organized according to which of the top distinguishing pathways for this class that they effect. Some genes affect multiple pathways, as indicated by the bars at left. The highlighted genes at right are all proteasomal subunits.</p>

### 2020-08-03 code/supplements.Rmd

>>>>

```{r, supplemental-s5-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

## class 7 and 8 distinguishing genes
genes <- c("PIK3CA", "KRAS", "TP53")

selected_mut <- gene_data %>% dplyr::select(external_gene_name, ensembl_gene_id) %>% filter(external_gene_name %in% genes) ## getting ensembl_id for the genes for filtering gdc data

s5_data <- umap_cancertype %>% dplyr::select(sample_id, plot_x, plot_y) %>% ## select relevant columns
  left_join(., selected_mut, by = "sample_id") %>%
  melt(data = ., id.vars = c("sample_id", "plot_x", "plot_y")) %>%
  filter(value != 0)

save(s5_data, file = "../data/s5_data.Rda")
```

```{r supplemental-s5-plot}
load("../data/s5_data.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

clust_178 <- umap_cancertype %>% filter(clust_knn %in% c(1, 7, 8))

umap_clust178 <- ggplot(clust_178, aes(x = plot_y, y = plot_x, colour = as.character(clust_knn))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") + labs(colour = "Cluster") +
  guides(colour = guide_legend(override.aes = list(size=3))) +
  scale_colour_manual(values = c("gold2", "gray51", "springgreen4")) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.ticks = element_blank(),
    legend.position = c(0.95, 0.8))

load("../data/s5_data.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
class7_8 <- umap_cancertype %>% filter(clust_knn %in% c(7, 8))
umap_bg_7_8 <- class7_8[,-"project_code"] ## get background data for facet

# pik3ca_kras <- ggplot(s5_data %>% filter(sample_id %in% umap_bg_7_8$sample_id) %>% filter(variable != "TP53") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
#   geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
#   geom_point(size = 0.25) +
#   theme_classic() +
#   coord_fixed() +
#   #ylim(-0.50, 1.5) +
#   xlab("UMAP 1") + ylab("UMAP 2") +
#   scale_colour_manual(values = rep("firebrick3", 20)) +
#   theme(axis.text.x = element_blank(),
#     axis.text.y = element_blank(),
#     axis.title.x = element_blank(),
#     axis.title.y = element_blank(),
#     axis.ticks = element_blank(),
#     axis.line = element_blank(),
#     legend.position = "none") +
#   facet_wrap(~ variable, ncol = 2)

tp53 <- ggplot(s5_data %>% filter(variable == "TP53") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ variable, ncol = 1)

pik3ca <- ggplot(s5_data %>%  filter(variable == "PIK3CA") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  #ylim(-0.50, 1.5) +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ variable, ncol = 1)

kras <- ggplot(s5_data %>%  filter(variable == "KRAS") , aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  #ylim(-0.50, 1.5) +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ variable, ncol = 1)

layout <- '
AABB
AACC
AADD
'

umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
```

### 2020-08-03 code/table1.Rmd

cluster_spec_enrich_table <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
save(cluster_spec_enrich_table, file = "../data/cluster_spec_enrich_table.Rda")
rm(list = ls(all.names = T))
gc()
```

```{r table-1}
load("../data/cluster_spec_enrich_table.Rda")
tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data

@@

cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>

>>>>

cluster_spec_enrich_table <- cluster_spec_enrich_table %>% do.call(rbind, .) %>% as.data.frame()
save(cluster_spec_enrich_table, file = "../data/cluster_spec_enrich_table.Rda")
rm(list = ls(all.names = T))
gc()
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1: Most frequently mutated gene in class-specific enriched pathways (effect size >= 0.30) for specific cancer**</p>
```{r table-1}
load("../data/cluster_spec_enrich_table.Rda")
tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data

@@

cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(format = "html", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```

### 2020-08-03 text/discussion.Rmd

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant clusters 8, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but are not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.

>>>>

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
mutant cluster 8, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify any cancers
that were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to histological
and molecular expression profiles that define them and how they respond to experimental
stimuli [@breast-subtypes2]. It has been hypothesized that differences in the
molecular regulators of development in the precursor cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 1J**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but are not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients
of a given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the
basis for some tumors being treatment-refractory in spite of receiving the
standard of clinical care for diagnostic markers remains elusive. Doubtless some
of this is due to bad luck, as tumors can metastasize and remain dormant years
before they are detected at distal sites, or resistant clones may have already
arisen at undetectable levels [@hallmarks_II], but our analysis suggests the
possibility of identifying more informative molecular, histological or cellular
subtypes that could form a basis for future separation of patients into
different precision treatment regimens.
